Meeting the Challenges of Modeling Astrophysical Thermonuclear Explosions: \castro, \maestro, and the \amrex\ Astrophysics Suite

### 2017-10-20 astronum_2017/astronum-amrex.tex
caption fix
some small edits

No single algorithm meets all of the demands imposed by these events.
Instead, we advance our understanding of these events by piecing
together simulations of different phases of the evolution from
different codes.  Here we discuss our simulation codes, \maestro\ and
\castro, designed to perform three-dimensional models of the early
subsonic evolution leading to runaway and the subsequent expolosion.
Together this suite of codes allows us to address many problems in
stellar and nuclear astrophysics.  We describe some of the design
details, the current architecture of the code, and some applications
below.

@@

Our interests are thermonuclear explosions, including Type Ia
supernovae (SNe Ia), X-ray bursts (XRBs), and novae.  The basic
ingredients for these events are thermonuclear energy release and a
degenerate equation of state that allows a runaway to build without a
pressure response.  Most of the current models for these events are
characterized by a long timescale ``simmmering'' phase where reactions
heat the star or layer and drive convection.  Eventually, reactions
become vigorous enough that a runaway takes place, perhaps with an
accompanying burning front that spreads through the star.

@@

There are open questions in all of these scenarios that can be
addressed through simulation.  For the Chandra and sub-Chandra models,
what is the distribution (spatial and temporal) of the hotspots setup
by turbulent convection that give rise to burning fronts?  A
longstanding question with the Chandra model is, can a deflagration
transition to a detonation during the burning front propagation
through the star.  For the sub-Chandra double detonation model, is it
possible to create a detonation in the thin surface He layer?  For
double degenerates, does the burning take place promptly? and can we
avoid an accretion-induced collapse to a neutron star.  For many of
these, we also need to address numerical issues, such as whether it is
possible to accurately model the ignition of a detonation with the
spatial resolution we can attain?  These are some of the questions we
seek to answer.

@@

Our suite of application codes is built on the
\amrex\ block-structured adaptive mesh refinement (AMR) library.  The
basic programming model has \amrex\ managing the grid data-structures
and parallel communications, and it calls the computational kernels on
a patch-by-patch basis.  The core library is written in \cpp\ with 
computational kernels written in Fortran\footnote{Currently
  \maestro\ is using an older pure-Fortran interface, but this will be
  ported to the \cpp\ \amrex\ this coming year.}---this allows us to take
advantage of the strengths of both languages. \amrex\ supports subcycling
in time, which we use in \castro.

\amrex\ uses a hybrid MPI + OpenMP approach to parallelism.
Distribution of grid patches to nodes using MPI provides a natural
coarse-grained approach to distributing the computational work, while
subdividing patches into tiles and using threads to parallelize over
tiles using OpenMP provides effective fine-scale parallelization and
amortizes thread overhead over large units of work.  For \castro\,
this tiling logically subdivides AMR patches to increase the
parallelism for OpenMP, without incurring a large increase in metadata
that would be associated with using smaller patches in a flat MPI
mode~\cite{tiling}.  This strategy is especially important for many
core architectures like the Intel Phi.  Ongoing development is being
done to support GPU offloading, using managed memory provided by the
latest generations of GPUs.

There are many application codes built on \amrex, including those in
combustion, multiphase flow, accelerator design, and microfluidics. In
astrophysics, these include \maestro\ and \castro\ for stellar and
nuclear astrophysics applications and \nyx~\cite{nyx} for cosmological
applications.  Here we focus on the former two.

@@

\maestro~\cite{MAESTRO:Multilevel} is a low Mach number stellar
hydrodynamics code designed for efficiently modeling convection in
stars.  \maestro\ decomposes the state into a one-dimensional
hydrostatic base state and a three-dimensional Cartesian state that
evolves the departures from HSE.  A constraint equation is derived by
requiring that the pressure everywhere be close to the background
hydrostatic state.  The constraint acts to enforce instantaneous
acoustic equilibriation, effectively filtering soundwaves from the
system, while retaining the compressibility effects due to the
background stratification of the star and local heat release, as well
as the hydrostatic adjustment of the star.  In this fashion, it is
more general than traditional anelastic methods.  The timestep
constraint for these equations depends only on the fluid velocity, not
the sound speed, enabling much larger timesteps than compressible
codes for highly subsonic flows.

@@

All of our simulation codes are open source and follow a fully open
development model---the development git repos are hosted on
github\footnote{\url{https://github.com/AMReX-Astro/}}, available for
anyone to see and contribute to using issues and pull-requests.
Additionally, we have mailing lists for discussions and asking for
help.  Several branches are used in our workflow.  New changes are put
into the {\tt development} branch in each repo.  Nightly regression
testing is used to ensure that no new bugs were introduced.  Once a
month, {\tt development} is merged into {\tt master}.  Finally, {\em all source
  files, model files, input parameters, etc.\ for any published
  science results are also available in the code repos.}  When
feasible, the git hashes for the published results are included in
paper acknowledgements.

@@

A key design goal of our application codes is performance portability.
We want the same kernels to run on clusters, manycore machines (like
Intel KNL), and GPU-based machines.  Our development has balanced this
need with architecture-specific optimizations to maximize code reuse.

@@

\begin{figure}[t]
\centering
\includegraphics[width=0.48\linewidth]{sedov_scaling}
\includegraphics[width=0.48\linewidth]{wdmerger_scaling}
\caption{\label{fig:castro-scaling} (left) \castro\ strong scaling on
  OLCF titan for a pure hydro (Sedov w/ real EOS) problem.  The
  different colors represent different base resolutions.  We see
  excellent strong and weak scaling for the single level runs.  For
  the $1024^3$ run, we also ran with one level of refinment by a
  factor of 2 (triangles) or a factor of 4 (squares), and see good
  strong scaling.  The variability (shown by the error bars) at high
  core counts shows we are becoming work-starved. (right)
  \castro\ strong scaling on OLCF titan for a hydro + Poisson gravity
  (wdmerger) problem.  This problem uses a multipole solver to
  determine Dirichlet boundary conditions representing an isolated
  mass distribution, and then geometric multigrid to solve for the
  potential in the interior.  Two coarse grid sizes are shown, and
  demonstrate great strong scaling.  We also look at a single level of
  refinement (by a factor of 4) on top of this coarse grid.}
\end{figure}

\begin{figure}[t]
\centering
\includegraphics[width=0.48\linewidth]{summitdev_scaling}
\includegraphics[width=0.48\linewidth]{cori_scaling}
\caption{\label{fig:knl-gpu-scaling} \castro\ scaling on new
  architectures.  The left figure shows the \castro\ GPU proxy app
  (\starlord) on OLCF summitdev.  We see good scaling behavior to 64
  GPUs in this test of a pure hydro problem (Sedov w/ real EOS).  On the
right, we explore performance on NERSC Cori (and Intel KNL platform).}
\end{figure}

Figure~\ref{fig:maestro_scaling} shows strong scaling for \maestro\ on
the 3-d XRB problem.  This is a typical
\maestro\ application~\cite{xrb3}, where burning can be a significant
part of the overall evolution.  We ran on both OLCF titan and NERSC
edison.  We see that the code scales well to $\mathcal{O}(10^4)$
processors.  The upturn at the end of the scaling curves on titan
reflect the change from 1 MPI task / 8 OpenMP threads per NUMA node to
1 MPI task / 16 OpenMP threads per compute node (2 NUMA nodes).  The
main limitation to the scaling at the moment is the multigrid solves
used to enforce the projection (in particular the nodal solver).
\maestro\ also does not yet take advantage of the tiling capabilities
offered in \amrex.  These will be addressed in the future as we port
\maestro\ to the \cpp\ \amrex\ codebase to take advantage of ongoing
development.

Figure~\ref{fig:castro-scaling} shows scaling behavior for a pure
hydro + real EOS problem and a hydro / real EOS, + self-gravity
problem.  All tests were run on OLCF titan.  The AMD processors in
titan can be used in an 8 or 16 cores / node configuration.  For all
these runs, we ran in the 16 cores / node mode.  All runs also used
the PGI 17.7 compilers.  We see excellent scaling behavior for the
pure hydro problem.  For the self-gravity problem, we scale well with
a uniform grid, but with a factor of 4 refinement for the AMR level on
top of this, we observe degraded performance at higher core counts.
This is a challenging problem, as only 3\% of the domain volume is
refined.  For a given core count, there are a number of different
combinations of MPI ranks and OpenMP threads we can use.  In general,
with multilevel problems, we found the best performance with fewer MPI
ranks and more OpenMP threads.

Our latest focus has been on GPU ports of our application codes.  A
small proxy app, \starlord, was created from \castro\ with just the
hydrodynamics and stellar equation of state.  It uses a simple
method-of-lines formulation of the hydrodynamics and advects 13
nuclear species in addition to the hydrodynamics.  To offload work on
GPUs, GPU support was added directly into \amrex\footnote{This support
  is currently on a feature branch in the git repo, awaiting merge
  into {\tt development}.}.  In \amrex, grids consist of a number of
boxes that enable domain decomposition for parallelism.  An iterator
loops over all of the boxes at the same level of refinement and passes
a data pointer into a Fortran kernel function where the work is
done. For the simulation state data that resides in each box, we have
modified the memory allocator so that it can use a CUDA allocator
(either device or managed memory).  The domain iterator is configured
to handle data motion to and from the device, so that the compute
kernels can operate on data that is presumed to already be there, and
the computation is decoupled from the memory management. Computation
on the data can then be performed with OpenACC, CUDA Fortran, or (more
recently) OpenMP 4.5. We have also built CUDA compute support into
\amrex\ so that a compute kernel can be transparently operated on
using CUDA Fortran without substantially modifying the kernels (and we
anticipate using a similar strategy to use OpenACC and/or OpenMP in
the future). This helps ensure that we can continue to maintain
performance portability in our simulation codes, by decoupling the
physics algorithms from the backend support used to implement them on
various compute architectures.\MarginPar{add CPU time for reference}

Figure~\ref{fig:knl-gpu-scaling} shows the performance of
\starlord\ on the summitdev platform at OLCF\footnote{summitdev
  consists of 2 IBM Power8 processors and 4 Nvidia Pascal GPUs per
  node.}.  The highest GPU count (128) corresponds to 32 nodes on
summitdev.  We see a nearly linear speedup with the number of GPUs,
indicating good weak scaling on this machine.  Efforts are underway to
complete the port of the GPU developments into \castro.  This figure
also shows our performance on the Intel KNL manycore processors (using
NERSC Cori).  This is for the same problem as the self-gravity test on
titan. \MarginPar{need more info: base grid?}  The KNL chips in Cori
have 68 cores than can be run with 1, 2, or 4 hardware threads each.
For all these runs, we ran with 4 threads / core (effectively 272
threads / chip).

\begin{figure}[t]
\centering
\includegraphics[width=0.48\linewidth]{CUDA-Fortran-VODE-Integrator}
\begin{minipage}[b]{0.48\linewidth}
\caption{\label{fig:cudaode} GPU speed up over single CPU core for two
  reaction networks.  The 13-isotope alpha-chain is a
  potential network for the sub-Chandra simulations we propose here.
  The 9-isotope Urca network is used for our Urca simulations.  A
  variety of grid sizes were used, $16^3$, $32^3$, $64^3$, and
  $128^3$.  In all cases, we see significant speed up on the GPU
  vs.\ a single core.  These speed-up numbers are from our testbed
  workstation with a Titan X Pascal board and Intel Xeon X5650 processors.}
\end{minipage}
\end{figure}

@@

Figure~\ref{fig:current-runs} shows some of our recent science
simulations.  The left panel is an image of convection in the helium
layer on a sub-Chandra white dwarf.  This is part of a study of the
early stages of the double detonation SNe Ia model.  Using \maestro,
we are able to model the convection in the He layer for many turnover
times and saw a range of outcomes depending on the mass of the white
dwarf and He layer, including a global He-nova-like behavior where the
entire layer runsway together and localized
burning~\cite{subchandra2}.  Work continues to characterize the ignition.

>>>>

No single algorithm meets all of the demands imposed by these events.
Instead, we advance our understanding of these events by piecing
together simulations of different phases of the evolution from
different codes.  Here we discuss our simulation codes, \maestro\ and
\castro, designed to perform three-dimensional models of the early
subsonic evolution leading to runaway and the subsequent explosion.
Together this suite of codes allows us to address many problems in
stellar and nuclear astrophysics.  We describe some of the design
details, the current architecture of the code, and some applications
below.

@@

Our interests are thermonuclear explosions, including Type Ia
supernovae (SNe Ia), X-ray bursts (XRBs), and novae.  The basic
ingredients for these events are thermonuclear energy release and a
degenerate equation of state that allows a runaway to build without a
pressure response.  Most of the current models for these events are
characterized by a long timescale ``simmering'' phase where reactions
heat the star or layer and drive convection.  Eventually, reactions
become vigorous enough that a runaway takes place, perhaps with an
accompanying burning front that spreads through the star.

@@

There are open questions in all of these scenarios that can be
addressed through simulation.  For the Chandra and sub-Chandra models,
what is the distribution (spatial and temporal) of the hotspots setup
by turbulent convection that give rise to burning fronts?  A
longstanding question with the Chandra model is, can a deflagration
transition to a detonation during the burning front propagation
through the star.  For the sub-Chandra double detonation model, is it
possible to create a detonation in the thin surface He layer?  For
double degenerates, does the burning take place promptly? Can we
avoid an accretion-induced collapse to a neutron star?  For many of
these investigations, we need to address numerical issues such as 
whether it is
possible to accurately model the ignition of a detonation with the
spatial resolution we can attain.  These are some of the questions we
seek to answer.

@@

Our suite of application codes is built on the
\amrex\ block-structured adaptive mesh refinement (AMR) library.  The
basic programming model has \amrex\ managing the grid data-structures
and parallel communications, and it calls the computational kernels on
a patch-by-patch basis.  The core library is written in \cpp\ with 
computational kernels written in Fortran\footnote{Currently
  \maestro\ is written in pure-Fortran, but will be
  ported to the updated \cpp\ \amrex\ framework this coming year.}---this allows us to take
advantage of the strengths of both languages. \amrex\ supports subcycling
in time, which we use in \castro.

\amrex\ uses a hybrid MPI + OpenMP approach to parallelism.
Distribution of grid patches to nodes using MPI provides a natural
coarse-grained approach to distributing the computational work, while
subdividing patches into tiles and using threads to parallelize over
tiles using OpenMP provides effective fine-scale parallelization and
amortizes thread overhead over large units of work.  Additionally,
tiling does not incur the large increase in metadata
associated with using smaller patches in a flat MPI
mode~\cite{tiling} since we can use fewer, larger patches.
This strategy is especially important for
many-core architectures like the Intel Phi.  Ongoing development is being
done to support GPU offloading, using managed memory provided by the
latest generations of GPUs.

There are many application codes built on \amrex, including those in
combustion, multiphase flow, accelerator design, and microfluidics. In
astrophysics, these include \maestro\ and \castro\ for stellar and
nuclear astrophysics applications and \nyx~\cite{nyx} for cosmological
applications.  Here we focus on the former two.
\MarginPar{ajn - I want to put a grid picture in this doc somewhere}

@@

\maestro~\cite{MAESTRO:Multilevel} is a low Mach number stellar
hydrodynamics code designed for efficiently modeling convection in
stars.  \maestro\ decomposes the state variables into a one-dimensional
hydrostatic base state and a three-dimensional Cartesian state that
models the deviation from hydrostatic equilibrium.
A constraint equation is derived by
requiring that the pressure everywhere be close to the background
hydrostatic pressure.  The constraint acts to enforce instantaneous
acoustic equilibration, effectively filtering soundwaves from the
system, while retaining the compressibility effects due to the
background stratification of the star and local heat release, as well
as the hydrostatic adjustment of the star.  In this fashion, it is
more general than traditional anelastic methods.  The timestep
constraint for these equations depends only on the fluid velocity, not
the sound speed, enabling much larger timesteps than compressible
codes for highly subsonic flows.

@@

All of our simulation codes are open source and follow a fully open
development model---the development git repos are hosted on
github\footnote{\url{https://github.com/AMReX-Astro/}}, available for
anyone to see and contribute to using issues and pull-requests.
Additionally, we have mailing lists for discussions and asking for
help.  Several branches are used in our workflow.  New changes are put
into the {\tt development} branch in each repo.  Nightly regression
testing is used to ensure that no new bugs were introduced.  Once a
month, {\tt development} is merged into {\tt master}.  Finally, {\em all source
  files, model files, input parameters, etc.\ for any published
  science results are also available in the code repos.}  When
feasible, the git hashes for the published results are included in
paper acknowledgments.

@@

A key design goal of our application codes is performance portability.
We want the same kernels to run on clusters, manycore machines (e.g.
Intel KNL), and GPU-based machines.  Our development has balanced this
need with architecture-specific optimizations to maximize code reuse.

@@

\begin{figure}[t]
\centering
\includegraphics[width=0.48\linewidth]{sedov_scaling}
\includegraphics[width=0.48\linewidth]{wdmerger_scaling}
\caption{\label{fig:castro-scaling} (left) \castro\ strong scaling on
  OLCF titan for a pure hydro (Sedov w/ real EOS) problem.  The
  different colors represent different base resolutions.  We see
  excellent strong and weak scaling for the single level runs.  For
  the $1024^3$ run, we also ran with one level of refinement by a
  factor of 2 (triangles) or a factor of 4 (squares), and see good
  strong scaling.  The variability (shown by the error bars) at high
  core counts shows we are becoming work-starved. (right)
  \castro\ strong scaling on OLCF titan for a hydro + Poisson gravity
  (wdmerger) problem.  This problem uses a multipole solver to
  determine Dirichlet boundary conditions representing an isolated
  mass distribution, and then geometric multigrid to solve for the
  potential in the interior.  Two coarse grid sizes are shown, and
  demonstrate great strong scaling.  We also look at a single level of
  refinement (by a factor of 4) on top of this coarse grid.}
\end{figure}

\begin{figure}[t]
\centering
\includegraphics[width=0.48\linewidth]{summitdev_scaling}
\includegraphics[width=0.48\linewidth]{cori_scaling}
\caption{\label{fig:knl-gpu-scaling} \castro\ scaling on new
  architectures.  The left figure shows the \castro\ GPU proxy app
  (\starlord) on OLCF summitdev.  We see good scaling behavior to 64
  GPUs in this test of a pure hydro problem (Sedov w/ real EOS).  On the
right, we explore performance on NERSC cori (and Intel KNL platform).}
\end{figure}

Figure~\ref{fig:maestro_scaling} shows strong scaling for \maestro\ on
the 3-d XRB problem.  This is a typical
\maestro\ application~\cite{xrb3}, where burning can be a significant
part of the overall evolution.  We ran on both OLCF titan and NERSC
edison.  We see that the code scales well to $\mathcal{O}(10^4)$
processors.  The upturn at the end of the scaling curves on titan
reflect the change from 1 MPI task / 8 OpenMP threads per NUMA node to
1 MPI task / 16 OpenMP threads per compute node (2 NUMA nodes).  The
main limitation to the scaling at the moment is the multigrid solves
used to enforce the projection (in particular the nodal solver).
Also, \maestro\ currently uses a simpler fine-grained approach to parallelism
where planes in the $z$-direction are divided among OpenMP threads.
As we port \maestro\ to the \cpp\ \amrex\ code base we will take
advantage of ongoing development for increased tile size control.

Figure~\ref{fig:castro-scaling} shows \castro\ scaling behavior for a pure
hydro + real EOS problem and a hydro / real EOS, + self-gravity
problem.  All tests were run on OLCF titan.  The AMD processors in
titan can be used in an 8 or 16 cores / node configuration.  For all
these runs, we ran in the 16 cores / node mode.  All runs also used
the PGI 17.7 compilers.  We see excellent scaling behavior for the
pure hydro problem.  For the self-gravity problem, we scale well with
a uniform grid, but with a factor of 4 refinement for the AMR level on
top of this, we observe degraded performance at higher core counts.
This is a challenging problem, as only 3\% of the domain volume is
refined.  For a given core count, there are a number of different
combinations of MPI ranks and OpenMP threads we can use.  In general,
with multilevel problems, we found the best performance with fewer MPI
ranks and more OpenMP threads.

Our latest focus has been on GPU ports of our application codes.  A
small proxy app, \starlord, was created from \castro\ with just the
hydrodynamics and stellar equation of state.  It uses a simple
method-of-lines formulation of the hydrodynamics and advects 13
nuclear species in addition to the hydrodynamics.  To offload work on
GPUs, GPU support was added directly into \amrex\footnote{This support
  is currently on a feature branch in the git repo, awaiting merge
  into {\tt development}.}.  In \amrex, grids consist of a number of
boxes that enable domain decomposition for parallelism.\MarginPar{unclear what the difference between a grid and box is here}
An iterator loops over all of the boxes at the same level of refinement and passes
a data pointer into a Fortran kernel function where the work is
done. For the simulation state data that resides in each box, we have
modified the memory allocator so that it can use a CUDA allocator
(either device or managed memory).  The domain iterator is configured
to handle data motion to and from the device, so that the compute
kernels can operate on data that is presumed to already be there, and
the computation is decoupled from the memory management. Computation
on the data can then be performed with OpenACC, CUDA Fortran, or (more
recently) OpenMP 4.5. We have also built CUDA compute support into
\amrex\ so that a compute kernel can be transparently operated on
using CUDA Fortran without substantially modifying the kernels (and we
anticipate using a similar strategy to use OpenACC and/or OpenMP in
the future). This helps ensure that we can continue to maintain
performance portability in our simulation codes, by decoupling the
physics algorithms from the backend support used to implement them on
various compute architectures.\MarginPar{add CPU time for reference}

Figure~\ref{fig:knl-gpu-scaling} shows the performance of
\starlord\ on the summitdev platform at OLCF\footnote{summitdev
  consists of 2 IBM Power8 processors and 4 Nvidia Pascal GPUs per
  node.}.  The highest GPU count (128) corresponds to 32 nodes on
summitdev.  We see a nearly linear speedup with the number of GPUs,
indicating good weak scaling on this machine.  Efforts are underway to
complete the port of the GPU developments into \castro.  This figure
also shows our performance on the Intel KNL manycore processors (using
NERSC cori).  This is for the same problem as the self-gravity test on
titan. \MarginPar{need more info: base grid?}  The KNL chips in cori
have 68 cores than can be run with 1, 2, or 4 hardware threads each.
For all these runs, we ran with 4 threads / core (effectively 272
threads / chip).

\begin{figure}[t]
\centering
\includegraphics[width=0.48\linewidth]{CUDA-Fortran-VODE-Integrator}
\begin{minipage}[b]{0.48\linewidth}
\caption{\label{fig:cudaode} GPU speed up over single CPU core for two
  reaction networks.  The 13-isotope alpha-chain is a
  potential network for the sub-Chandra simulations.
  The 9-isotope Urca network is used for our Urca simulations.  A
  variety of grid sizes were used, $16^3$, $32^3$, $64^3$, and
  $128^3$.  In all cases, we see significant speed up on the GPU
  vs.\ a single core.  These speed-up numbers are from our testbed
  workstation with a Titan X Pascal board and Intel Xeon X5650 processors.}
\end{minipage}
\end{figure}

@@

Figure~\ref{fig:current-runs} shows some of our recent science
simulations.  The left panel is an image of convection in the helium
layer on a sub-Chandra white dwarf.  This is part of a study of the
early stages of the double detonation SNe Ia model.  Using \maestro,
we are able to model the convection in the He layer for many turnover
times and saw a range of outcomes depending on the mass of the white
dwarf and He layer, including a global He-nova-like behavior where the
entire layer runs way together and localized
burning~\cite{subchandra2}.\MarginPar{incomplete sentence}
We are performing further studies to characterize the ignition.
